---
### YAML section (use # for comments)
title: "Sc-genetics: RolyPoly main"
author: "Pascal N. Timshel, Pers Lab"
date: "`r Sys.time()`" # Sys.Date()
output:
  html_notebook: 
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---


```{r}
### Extract the code from the notebook
# knitr::purl("rolypoly_main.Rmd", output = "rolypoly_main.R", documentation = 2)
```


# Setup

```{r, message = FALSE, results='hide', warning=FALSE}
library(rolypoly)
library(tidyverse)
library(ggplot2)
```

```{r}
wd <- "/projects/timshel/sc-genetics/sc-genetics/src/RP_BMI/"
setwd(wd)

dir.sc_genetics.data <- "/projects/timshel/sc-genetics/sc-genetics/data" # no trailing slash

```


# PARAMETERS

```{r}
dir.ldfiles <- file.path(dir.sc_genetics.data, "rolypoly/EUR_LD_FILTERED_NONAN_R") # Linkage disequilibrium files
```


# GWAS

REQUIRED COLUMNS IN DATA FRAME: c('chrom', 'pos', 'rsid', 'beta', 'se', 'maf')

```{r}
# file.gwas <- file.path(dir.sc_genetics.data, "gwas_sumstats/body_BMI_Locke2015.gwassumstats.rolypoly_fmt.tab.gz") # BMI
file.gwas <- file.path(dir.sc_genetics.data, "gwas_sumstats/mental_SCZ_Ripke2014.gwassumstats.rolypoly_fmt.tab.gz") # SCZ
df.gwas <- read_delim(file.gwas, delim="\t") # suppressMessages or use "col_types=cols()"

# TODO: check that chr is always integer
```

```{r}
### Number of MHC/HLA SNPs [chr6:25Mb-34Mb]
df.gwas %>% filter( (((chr == 6)&(pos >= 25e6)) & ((chr == 6)&(pos <= 34e6))) ) %>% nrow() # --> 13852 HLA SNPs for BMI.
```

## Format and filter GWAS: exclude HLA

Rename columns
```{r}
df.gwas.rp <- df.gwas %>% 
  rename(
    rsid = rsID,
    chrom = chr,
    maf = snp_maf) %>%
  filter( !(((chrom == 6)&(pos >= 25e6)) & ((chrom == 6)&(pos <= 34e6))) ) # remove MHC/HLA [chr6:25Mb-34Mb]
  
# must be data frame for RolyPoly to run

df.gwas.rp %>% head
```



# Gene annotation

```{r}
file.gene_annot <- file.path(dir.sc_genetics.data, "gene_annotations/gene_annotation.hsapiens_mmusculus_unique_orthologs.GRCh37.ens_v91.txt.gz")
df.gene_annot <- read_delim(file.gene_annot, col_types=cols(), delim="\t") # suppressMessages or use "col_types=cols()"
```


```{r}
df.gene_annot %>% head
```

## Filter genes: filter out non-autosomal genes (discard X, Y, contigs etc)
```{r}
n.discarded <- df.gene_annot %>% filter(!chromosome_name %in% 1:22) %>% nrow()
print(sprintf("Number of non-autosomal genes discarded: %s", n.discarded))

df.gene_annot.filtered <- df.gene_annot %>% filter(chromosome_name %in% 1:22)
```


# Expression data

REQUIRED COLUMNS IN DATA FRAME: none (just cell-type/tissue names); rownames is genes

```{r}
### t-test data
# file.expr <- file.path(dir.sc_genetics.data, "expression/arc_lira/arc_lira.celltype_expr.ttest.hsapiens_orthologs.csv.gz") # arc_lira
# file.expr <- file.path(dir.sc_genetics.data, "expression/maca/maca.per_tissue.celltype_expr.ttest.hsapiens_orthologs.csv.gz") # maca per_tissue


### average expression
# file.expr <- file.path(dir.sc_genetics.data, "expression/maca/maca.per_tissue.celltype_expr.avg_expr.hsapiens_orthologs.csv.gz") # maca per_tissue
file.expr <- file.path(dir.sc_genetics.data, "expression/maca/maca.per_tissue_celltype.celltype_expr.avg_expr.hsapiens_orthologs.csv.gz") # maca per_tissue_celltype

df.expr <- read_csv(file.expr)
```

```{r}
df.expr %>% head
```


## Format for RolyPoly and filter genes (based on `df.gene_annot.filtered`, e.g. autosomal)


```{r}
bool.expr_genes_filtered <- df.expr$gene %in% df.gene_annot.filtered$ensembl_gene_id
df.expr.filtered <- df.expr %>% 
  filter(bool.expr_genes_filtered) %>%
  as.data.frame() %>%
  column_to_rownames(var="gene")
head(df.expr.filtered)
```


# Numerical transformation of expression data (square values)

1. [only for non-ttest value]: Z-score (R scale)
    1. z-scores for each gene (across cells) "cell’s relative expression of each gene, in comparison to all other cells:
2. Square to avoid negative values.

```{r}
### Square values
df.expr.filtered.rp <- df.expr.filtered^2
# df.expr.filtered.rp <- abs(df.expr.filtered)
```



# Block annotation: window size, etc

REQUIRED COLUMNS IN DATA FRAME: c('chrom', 'start', 'end', 'label')


```{r}
WINDOW_SIZE <- 10e3 # 10kb

df.block_annotation <- df.gene_annot.filtered %>% transmute(
  label=ensembl_gene_id,
  chrom=chromosome_name,
  start=start_position-WINDOW_SIZE, # TSS-WINDOW_SIZE
  end=start_position+WINDOW_SIZE # TSS+WINDOW_SIZE
)

df.block_annotation %>% head
```


# Rolling rolypoly

We include all the previously described data into the main rolypoly function call. This function has many parameters to tinker with, however, should run fine with the defaults. Most importantly consider the number of bootstrap iterations to get accurate standard errors. We usually use at least 200.

```{r}

df.gwas.rp.test <- df.gwas.rp %>% slice(1:1000)


```






```{r}
library(doParallel)
registerDoParallel(40) # half of the number of cores 
getDoParWorkers()
```



```{r, results='hide'}
# message = FALSE

# TODO: run with "highres" data

rm(rp)
system.time(rp <- rolypoly_roll(
  gwas_data = as.data.frame(df.gwas.rp), # df.gwas.rp.test
  block_annotation = as.data.frame(df.block_annotation),
  block_data = as.data.frame (df.expr.filtered.rp),
  ld_folder = dir.ldfiles,
  bootstrap_iters = 100,
  gwas_link_parallel = T,
  bootstrap_parallel = T
))
```


```{r}
# rp.scz.maca.tissue <- rp
# save(rp.scz.maca.tissue, file="out.rolypoly_objs.SCZ.RData")

# rp.arc_lira <- rp
# rp.maca.tissue <- rp

# save(rp.arc_lira, rp.maca.tissue, file="out.rolypoly_objs.BMI.RData")
# load(file="out.rolypoly_objs.BMI.RData")

# save.image(file="out.rolypoly_main-NB.RData")
# load(file="out.rolypoly_main-SCRIPT.RData")
```


# Parallelization: Inference on other expression data set

Without any parallelization rolypoly could take a couple hours to run. Much of the computation is spent linking SNPs to genes and reading in LD information. If one provides the a rolypoly object to the rolypoly parameter of the `rolypoly_roll` function call and a new object of block data, then only inference is performed. For example with the previous rolypoly object we would run inference on a new expression data set with the following call (assuming that the gene labels did not change between datasets).

```{r}
# OBS: assuming that the gene labels did not change between datasets.
# rp.new <- rolypoly_roll(
#   # some new set of expression data
#   rolypoly = rp,
#   block_data = new_sim_expression_data_normalized
# )
```

Thus, one can run precomputation (slow) once and then rerun inference (faster) on various expression matrices more quickly.


## Results

Once rolypoly is finished we can access all the results within the returned rolypoly object. We generated the GWAS under the model with effects of 0.02 and 0.01 for the Liver and Blood tissues. To take a look at the inferred parameters use:

```{r}
rp$full_results$parameters %>% sort
```

Better yet, there's a data frame with results from the bootstrap runs that calculates standard errors for these paramter estimates, p-values, and 95\% confidence intervals.

```{r}
rp$bootstrap_results %>% arrange(-bt_value) %>% head
```

For visualization, one can use the following function which plots the estimate with 95\% confidence intervals,

```{r, fig.show='asis', fig.width=7, fig.height=8}
plot_rolypoly_annotation_estimates(rp)
```

Additionally, we plot the $-log10(p)$ to rank tissues by the strength of their association

```{r, fig.show='asis', fig.width=7, fig.height=8}
plot_rolypoly_annotation_ranking(rp)
```

These functions return ggplot2 objects so free to manipulate them as such.









```{r}
########################################################################################################################
########################################################################################################################
################################################## LEFTOVER ############################################################
```


Check that all genes are found in the gene annotation file
Ortherwise we would have to filter out genes not in the gene annotation file...


















